---
title: "Prepare_data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(impIndicator)
```

# Process GBIF data from the environment
1. import GBIF data using `read.csv()`, `readr::read_csv()`, or `readxl::read_excel()` based on the data set format.

Here is an example a GBIF occurrences data with required columns.`decimalLatitude`,
`decimalLongitude`,`species`,`speciesKey`,`coordinateUncertaintyInMeters` and
`year`
```{r echo=FALSE, results='asis'}
library(knitr)
taxa_Acacia %>% dplyr::select(decimalLatitude,
                      decimalLongitude,
                      species,
                      speciesKey,
                      coordinateUncertaintyInMeters,
                      year
) %>% head(10) %>% 
  kable()
```

The `region` of the study has to be given as a shapefile.

```{r}
acacia_cube <- taxa_cube(
  taxa = taxa_Acacia,
  region = southAfrica_sf,
  res = 0.25,
  first_year = 2010,
  last_year = 2023
)

```

The cube is a `sim_cube` object. Below is an the example of acacia taxa
```{r}
# view processed cube
acacia_cube$cube

# view coordinates
head(acacia_cube$coords)
```
