---
title: "Prepare data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(impIndicator)

library(b3gbi) # General biodiversity indicators for data cubes
library(dplyr) # Data wrangling
library(knitr) # Nice tables
```

## Process GBIF data from the R environment
1. import GBIF data using `read.csv()`, `readr::read_csv()`, or `readxl::read_excel()` based on the data set format.

Here is an example a GBIF occurrences data with the minimum required columns.`decimalLatitude`, `decimalLongitude`, `species`, `speciesKey`, `coordinateUncertaintyInMeters` and `year`

```{r echo=FALSE, results='asis'}
taxa_Acacia %>%
  dplyr::select(
    "decimalLatitude",
    "decimalLongitude",
    "species",
    "speciesKey",
    "coordinateUncertaintyInMeters",
    "year"
  ) %>%
  head(10) %>%
  kable()
```

The `region` of the study has to be given as a shapefile of the study area 
or a character representing the country of study area . An example is:

```{r}
southAfrica_sf
```

```{r}
acacia_cube <- taxa_cube(
  taxa = taxa_Acacia,
  region = southAfrica_sf,
  first_year = 2010,
  last_year = 2023
)
```

The cube is a `sim_cube` object. Below is an example of the acacia taxa in 
South Africa
```{r}
# View processed cube
acacia_cube
```
## Download from GBIF website
The cube can be generated by downloading the GBIF with `rgbif::occ_data()` 

## Cube with standard grid
**impIndicator** works other cubes with standard grid cell, such as, eea and eqdgc
which are processed from `b3gbi::process_cube()`. An example is the mammal_cube
in the `b3gbi` package.

```{r}
# Load GBIF data cube
cube_name <- system.file("extdata", "denmark_mammals_cube_eqdgc.csv",
                         package = "b3gbi")

# Prepare cube
mammal_cube <- b3gbi::process_cube(cube_name, first_year = 2000)

# View cube
mammal_cube
```
